\textbf{ Interpretable Gaussian Processes for Stellar Light Curves }

### 2020-11-13 tex/gitlinks.tex
more text

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/0f5cd85041405565c68a93eca39244838420c99d/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/0f5cd85041405565c68a93eca39244838420c99d/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/0f5cd85041405565c68a93eca39244838420c99d/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

>>>>

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/3c024c44d780c51442e91d488e120a8eb4a05da5/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/3c024c44d780c51442e91d488e120a8eb4a05da5/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/3c024c44d780c51442e91d488e120a8eb4a05da5/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

### 2020-11-13 tex/ms.tex
more text

    \mathbf{x} = \left( c \,\, \lambda \,\, \phi \,\, r \right)^\top

@@

    c

@@

    \mathbf{R}_{\hat{\mathbf{y}}}(\lambda)

@@

    \mathbf{R}_{\hat{\mathbf{x}}}(\phi)

@@

    \mathbf{s}(r)

@@

where $c$ is the contrast of a spot,
$\lambda$ is its longitude, $\phi$ is its latitude,
and $r$ is its radius.
The vector function $\mathbf{s}(r)$

@@

circular spot at $\lambda = \phi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\phi)$ is the Wigner matrix that rotates the

@@

latitude $\phi$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\lambda)$ is the Wigner

@@

longitude $\lambda$; these three functions are detailed in the sections below.

@@

$\mathbf{x}$ describing the size, brightness, and position of a spot to the

@@

is separable in each of the four spot properties:

@@

    p(\phi \, \big| \, \pmb{\theta}_{\phi})\,
    p(r \, \big| \, \pmb{\theta}_{r})

@@

@@

@@

@@

We devote the next several sections to the computation of these eight
integrals.

@@

\subsection{Spot profile}
%
We model the brightness $b$ an angle $\vartheta$ away from the
center of a spot of negative unit intensity and radius $r$ be
%
\begin{align}
    b(r; \vartheta) & = \frac{1}{1 + \exp\left(\dfrac{r-\vartheta}{s}\right)} - 1
\end{align}
%
for some (constant) shape parameter $s$. In the limit $s \rightarrow 0$, $b$ approaches an
inverted top-hat function with half-width equal to $r$,
corresponding to a circular spot of uniform intensity. For $s > 0$, each half
of $b$ is a sigmoid with half-width at half-minimum equal to $r$.
In our implementation of the algorithm we choose $s = 0.2^\circ$,
which is small compared to features of interest but not too small as to
create numerical issues when computing model gradients (which would be
undefined at the spot boundary if the spot profile were truly an inverted top-hat).

Our goal now is to expand the function above in spherical harmonics. To that end,
we note that in a frame where the spot is centered on $\hat{\mathbf{z}}$
(i.e., at polar angle $\vartheta = 0$), the brightness profile is azimuthally
symmetric, so the only nonzero coefficients in the spherical harmonic
expansion are those with order $m = 0$. The corresponding spherical harmonics are
simply proportional to the Legendre polynomials in $\cos\vartheta$, so our task is
simplified to finding the Legendre polynomial expansion of $b$.
Define a vector $\pmb{\vartheta}$ of $K$ equally-spaced points
between $0$ and $\pi$, with coefficients given by
%
\begin{align}
    \vartheta_k = \frac{k\pi}{K-1}
    \quad.
\end{align}
%
We wish to model the brightness evaluated at each $\vartheta_k$
as a weighted combination of Legendre polynomials,
%
\begin{align}
    \mathbf{b}(r) = \mathbf{B} \, \tilde{\mathbf{s}}(r)
\end{align}
%
%
where $\mathbf{B}$ is a design matrix whose columns are
the weighted Legendre polynomials,
%
\begin{align}
    B_{k,l} & = \sqrt{2l + 1} \, P_l\left( \cos \vartheta_k \right)
    \quad,
\end{align}
%
and $\tilde{\mathbf{s}}(r)$ are the coefficients of the expansion.
These are related to the full vector of spherical harmonic coefficients describing
the spot, $\mathbf{s}(r)$, by
%
\begin{align}
    s_{l,m}(r) = \tilde{s}_l(r) \delta_{m,0}
    \quad,
\end{align}
%
where $\delta$ is the Kronecker delta function.
To find the coefficients $\tilde{\mathbf{s}}(r)$, we solve the (linear) inverse problem,
%
\begin{align}
    \tilde{\mathbf{s}}(r) & = \mathbf{B}^+ \, \mathbf{b}(r)
\end{align}
%
%
where
%
\begin{align}
    \mathbf{B}^+ & = \mathbf{S} \Big(\mathbf{B}^\top \mathbf{B} +
    \epsilon \mathbf{I}\Big)^{-1} \mathbf{B}^\top
\end{align}
%
is the smoothed pseudo-inverse of $\mathbf{B}$ with small regularization
parameter $\epsilon$, $\mathbf{I}$ is the identity matrix, and
$\mathbf{S}$ is a diagonal smoothing matrix with coefficients
%
\begin{align}
    S_{k,l} = \exp\left[-\frac{l(l + 1)}{2\sigma^2}\right] \delta_{k,l}
\end{align}
%
for smoothing strength $\sigma$. For $\epsilon \rightarrow 0$ and
$\sigma \rightarrow \infty$,
$\mathbf{B}^+$ is the exact pseudo-inverse of
$\mathbf{B}$. However,
$\epsilon > 0$ is chosen for improved numerical stability and
$\sigma > 0$ is chosen to mitigate the effect of ringing in the solution.
In practice, we obtain good results with $\epsilon \approx 10^{-9}$
and $\sigma \approx 15$.

@@

Figure~\ref{fig:spot_profile} shows the intensity profile for spots of
different radii expanded to spherical harmonic degree $l_{\mathrm{max}} = \LMAX$,
computed from
%
\begin{align}
    \mathbf{b'}(r) & = \mathbf{B}
    \mathbf{B}^+ \, \mathbf{b}(r) \quad.
\end{align}
%
The average intensity
within the spots is close to $-1$ and the half-widths at half-minimum are
equal to the spot radii, as expected.
The effect of ringing due to the truncated spherical harmonic expansion is
evident, although it is strongly suppressed compared to an expansion
without the smoothing term (i.e., $\sigma = \infty$). However, for
$r \lesssim 10^\circ$, an expansion to $l_{\mathrm{max}} = \LMAX$ is
insufficient to correctly model the spot, as can be seen from the
$r = 5^\circ$ profile (dashed curve). Expansions to higher spherical harmonic
degree allow one to model spots with radii smaller than $10^\circ$, although
at increased computational cost and potential numerical stability issues;
we discuss this point at length in \S\ref{sec:tiny-spots}.

@@

\subsection{First moment}
%
The first moment of the radius distribution is (Equation~\ref{eq:e1})
%
\begin{align}
    \mathbf{e}_r
     & \equiv
    \int
    \mathbf{s}(r) \,
    p(r \, \big| \, \pmb{\theta}_{r}) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{1}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \mathbf{s}(r) \,
    \mathrm{d}r
    \quad.
\end{align}
%
Using the equations from the previous section, its components may be written
%
\begin{align}
    {e_r}_{l,m}
    &=
    \frac{\delta_{m,0}}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \tilde{s}_{l,m}(r) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0}}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,m,k}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    b_{k}(r)
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0} s}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,m,k}
    \ln 
    \left(
        \frac{
            1+\exp\left(\frac{r -\Delta -\vartheta_k}{s}\right)
        }
        {
            1+\exp\left(\frac{r + \Delta -\vartheta_k}{s}\right)
        }
    \right)
    \quad.
\end{align}
%
In the limit $\Delta \rightarrow 0$,
%
\begin{align}
    \lim_{\Delta \rightarrow 0}
    \mathbf{e}_r
    &=
    \mathbf{B}^+ \mathbf{b}(r)
    \quad.
\end{align}
%

\subsection{Second moment}
%
The second moment of the radius distribution is (Equation~\ref{eq:E1})
%
\begin{align}
    \mathbf{E}_r
     & \equiv
    \int
    \mathbf{s}(r) \,
    \mathbf{s}^\top(r) \,
    p(r \, \big| \, \pmb{\theta}_{r}) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{1}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \mathbf{s}(r) \,
    \mathbf{s}^\top(r) \,
    \mathrm{d}r
    \quad.
\end{align}
%
As before, its components may be written
%
%
\begin{align}
    {E_r}_{l,m,l',m'}
    &=
    \frac{\delta_{m,0}\delta_{m',0}}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \tilde{s}_{l,m}(r) \,
    \tilde{s}_{l',m'}(r) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0}\delta_{m',0}}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,m,k}
    \sum_{k'=0}^{K-1} B^+_{l',m',k'}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    b_{k}(r)
    b_{k'}(r)
    \mathrm{d}r
    \nonumber \\
    &= 
    \cdots
\end{align}
\xxx{...}

>>>>

    \mathbf{x} = 
    \left( 
        c_0 \,\, \cdots \,\, c_{N-1} \,\, 
        ,
        \lambda_0 \,\, \cdots \,\, \lambda_{N-1} \,\, 
        ,
        \phi_0 \,\, \cdots \,\, \phi_{N-1} \,\, 
        ,
        r_0 \,\, \cdots \,\, r_{N-1} \,\, 
    \right)^\top

@@

    \sum_{n=0}^{N-1}
    c_n

@@

    \mathbf{R}_{\hat{\mathbf{y}}}(\lambda_n)

@@

    \mathbf{R}_{\hat{\mathbf{x}}}(\phi_n)

@@

    \mathbf{s}(r_n)

@@

where $c_n$ is the contrast of the $n^\mathrm{th}$ spot,
$\lambda_n$ is its longitude, $\phi_n$ is its latitude,
$r_n$ is its radius, and $N$ is the total number of spots.
The vector function $\mathbf{s}(r_n)$

@@

circular spot of radius $r_n$ at $\lambda = \phi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\phi_n)$ is the Wigner matrix that rotates the

@@

latitude $\phi_n$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\lambda_n)$ is the Wigner

@@

longitude $\lambda_n$; these three functions are detailed in the sections below.

@@

$\mathbf{x}$ describing the size, brightness, and position of spots to the

@@

is separable in each of the four spot properties, and that all of the spots
are drawn from the same distribution:

@@

    \prod_{n=0}^{N-1}
    p(c_n \, \big| \, \pmb{\theta}_{c}) \,
    p(\lambda_n \, \big| \, \pmb{\theta}_{\lambda}) \,
    p(\phi_n \, \big| \, \pmb{\theta}_{\phi})\,
    p(r_n \, \big| \, \pmb{\theta}_{r})

@@

is the vector of yet-to-be-specified hyperparameters describing the process.

@@

    N \mathbf{e}_c

@@

    N \mathbf{E}_c

@@

In Equations~(\ref{eq:exp_y_sep}) and (\ref{eq:exp_yy_sep}), we used the fact 
that both the mean and the variance of the sum of $N$
independent, identically-distributed random variables are equal to $N$
times the individual mean and the variance, respectively.

We devote the next several sections to the computation of these eight
integrals.

@@

\subsection{Spot profile}
%
We model the brightness $b$ an angle $\vartheta$ away from the
center of a spot of negative unit intensity and radius $r$ be
%
\begin{align}
    \label{eq:brvartheta}
    b(r; \vartheta) & = \frac{1}{1 + \exp\left(\dfrac{r-\vartheta}{s}\right)} - 1
\end{align}
%
for some (constant) shape parameter $s$. In the limit $s \rightarrow 0$, $b$ approaches an
inverted top-hat function with half-width equal to $r$,
corresponding to a circular spot of uniform intensity. For $s > 0$, each half
of $b$ is a sigmoid with half-width at half-minimum equal to $r$.
In our implementation of the algorithm we choose $s = 0.2^\circ$,
which is small compared to features of interest but not too small as to
create numerical issues when computing model gradients (which would be
undefined at the spot boundary if the spot profile were truly an inverted top-hat).

Our goal now is to expand the function above in spherical harmonics. To that end,
we note that in a frame where the spot is centered on $\hat{\mathbf{z}}$
(i.e., at polar angle $\vartheta = 0$), the brightness profile is azimuthally
symmetric, so the only nonzero coefficients in the spherical harmonic
expansion are those with order $m = 0$. The corresponding spherical harmonics are
simply proportional to the Legendre polynomials in $\cos\vartheta$, so our task is
simplified to finding the Legendre polynomial expansion of $b$.
Define a vector $\pmb{\vartheta}$ of $K$ equally-spaced points
between $0$ and $\pi$, with coefficients given by
%
\begin{align}
    \vartheta_k = \frac{k\pi}{K-1}
    \quad.
\end{align}
%
We wish to model the brightness evaluated at each $\vartheta_k$
as a weighted combination of Legendre polynomials,
%
\begin{align}
    \mathbf{B} \, \tilde{\mathbf{s}}(r) = \mathbf{b}(r)
\end{align}
%
%
where $\mathbf{b}(r)$ is computed by evaluating Equation~(\ref{eq:brvartheta})
at each of the $\vartheta_k$,
$\mathbf{B}$ is a design matrix whose columns are
the weighted Legendre polynomials,
%
\begin{align}
    B_{k,l} & = \sqrt{2l + 1} \, P_l\left( \cos \vartheta_k \right)
    \quad,
\end{align}
%
and $\tilde{\mathbf{s}}(r)$ are the coefficients of the expansion.
These are related to the full vector of spherical harmonic coefficients describing
the spot, $\mathbf{s}(r)$, by
%
\begin{align}
    s^l_m(r) = \tilde{s}_l(r) \delta_{m,0}
    \quad,
\end{align}
%
where $\delta$ is the Kronecker delta function.
To find the coefficients $\tilde{\mathbf{s}}(r)$, we solve the (linear) inverse problem,
%
\begin{align}
    \tilde{\mathbf{s}}(r) & = \mathbf{B}^+ \, \mathbf{b}(r)
\end{align}
%
%
where
%
\begin{align}
    \mathbf{B}^+ & = \mathbf{S} \Big(\mathbf{B}^\top \mathbf{B} +
    \epsilon \mathbf{I}\Big)^{-1} \mathbf{B}^\top
\end{align}
%
is the smoothed pseudo-inverse of $\mathbf{B}$ with small regularization
parameter $\epsilon$, $\mathbf{I}$ is the identity matrix, and
$\mathbf{S}$ is a diagonal smoothing matrix with coefficients
%
\begin{align}
    S_{k,l} = \exp\left[-\frac{l(l + 1)}{2\sigma^2}\right] \delta_{k,l}
\end{align}
%
for smoothing strength $\sigma$. For $\epsilon \rightarrow 0$ and
$\sigma \rightarrow \infty$,
$\mathbf{B}^+$ is the exact pseudo-inverse of
$\mathbf{B}$. However,
$\epsilon > 0$ is chosen for improved numerical stability and
$\sigma > 0$ is chosen to mitigate the effect of ringing in the solution.
In practice, we obtain good results with $\epsilon \approx 10^{-9}$
and $\sigma \approx 15$.

@@

Figure~\ref{fig:spot_profile} shows the intensity profile for spots of
different radii expanded to spherical harmonic degree $l_{\mathrm{max}} = \LMAX$.
% computed from
% %
% \begin{align}
%     \mathbf{b'}(r) & = \mathbf{B}
%     \mathbf{B}^+ \, \mathbf{b}(r) \quad.
% \end{align}
%
The average intensity
within the spots is close to $-1$ and the half-widths at half-minimum are
equal to the spot radii, as expected.
The effect of ringing due to the truncated spherical harmonic expansion is
evident, although it is strongly suppressed compared to an expansion
without the smoothing term (i.e., $\sigma = \infty$). However, for
$r \lesssim 10^\circ$, an expansion to $l_{\mathrm{max}} = \LMAX$ is
insufficient to correctly model the spot, as can be seen from the
$r = 5^\circ$ profile (dashed curve). Expansions to higher spherical harmonic
degree allow one to model spots with radii smaller than $10^\circ$, although
at increased computational cost and potential numerical stability issues;
we discuss this point at length in \S\ref{sec:tiny-spots}.

@@

\subsection{First moment}
%
The first moment of the radius distribution is (Equation~\ref{eq:e1})
%
\begin{align}
    \mathbf{e}_r
     & \equiv
    \int
    \mathbf{s}(r) \,
    p(r \, \big| \, \pmb{\theta}_{r}) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{1}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \mathbf{s}(r) \,
    \mathrm{d}r
    \quad.
\end{align}
%
Using the equations from the previous section, its components may be written
%
\begin{align}
    (e_r)^l_m
    &=
    \frac{\delta_{m,0}}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \tilde{s}_{l}(r) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0}}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,k}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    b_{k}(r)
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0} s}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,k}
    \ln 
    \left(
        \frac{
            1+\exp\left(\frac{r -\Delta -\vartheta_k}{s}\right)
        }
        {
            1+\exp\left(\frac{r + \Delta -\vartheta_k}{s}\right)
        }
    \right)
    \quad.
\end{align}
%
In the limit $\Delta \rightarrow 0$,
%
\begin{align}
    \lim_{\Delta \rightarrow 0}
    \mathbf{e}_r
    &=
    \mathbf{B}^+ \mathbf{b}(r)
    \quad.
\end{align}
%

\subsection{Second moment}
%
The second moment of the radius distribution is (Equation~\ref{eq:E1})
%
\begin{align}
    \mathbf{E}_r
     & \equiv
    \int
    \mathbf{s}(r) \,
    \mathbf{s}^\top(r) \,
    p(r \, \big| \, \pmb{\theta}_{r}) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{1}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \mathbf{s}(r) \,
    \mathbf{s}^\top(r) \,
    \mathrm{d}r
    \quad.
\end{align}
%
As before, its components may be written
%
%
\begin{align}
    (E_r)^{l,l'}_{m,m'}
    &=
    \frac{\delta_{m,0}\delta_{m',0}}{2\Delta}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    \tilde{s}^l_{m}(r) \,
    \tilde{s}^{l'}_{m'}(r) \,
    \mathrm{d}r
    \nonumber \\
    &=
    \frac{\delta_{m,0}\delta_{m',0}}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,k}
    \sum_{k'=0}^{K-1} B^+_{l',k'}
    \int_{r_0 - \Delta}^{r_0 + \Delta}
    b_{k}(r)
    b_{k'}(r)
    \mathrm{d}r
    \nonumber \\
    &= 
    \frac{\delta_{m,0}\delta_{m',0}s}{2\Delta}
    \sum_{k=0}^{K-1} B^+_{l,k}
    \sum_{k'=0}^{K-1} B^+_{l',k'}
    C_{k,k'}(r)
    \quad,
\end{align}
%
where
%
\begin{align}
    C_{k,k'}(r)
    &\equiv
    \frac{
        \exp\left(\frac{\vartheta_{k} - \vartheta_{k'}}{s}\right)
        \ln 
        \left(
            \frac{
                1+\exp\left(\frac{r -\Delta -\vartheta_{k}}{s}\right)
            }
            {
                1+\exp\left(\frac{r + \Delta -\vartheta_{k}}{s}\right)
            }
        \right)
        -
        \ln 
        \left(
            \frac{
                1+\exp\left(\frac{r -\Delta -\vartheta_{k'}}{s}\right)
            }
            {
                1+\exp\left(\frac{r + \Delta -\vartheta_{k'}}{s}\right)
            }
        \right)
    }{
        1
        -
        \exp\left(\frac{\vartheta_{k} - \vartheta_{k'}}{s}\right)
    }
    \quad.
\end{align}
%
In the limit $\Delta \rightarrow 0$,
%
\begin{align}
    \lim_{\Delta \rightarrow 0}
    \mathbf{E}_r
    &=
    \mathbf{B}^+ \mathbf{b}(r) \mathbf{b}^\top(r) {\mathbf{B}^+}^\top
    \quad.
\end{align}
%
